Interfacing logic programming and Coq

### 2020-10-27 papers/types2020/sec3.tex
after main merge

{\color{red} The goal in this section is to say a bit about focusing
  generally, and to give a bit of detail about FPC for Horn clauses.
  Initial draft of text taken from \cite{blanco19ppdp}.
}

\begin{metanote}
  (AM): I would start with the logical reconstruction of \texttt{check}, adding the LJT-style proof terms~c\cite{Herbelin94,LengrandDM06} to fig \ref{fig:augmented} and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.
\end{metanote}
\subsection{Focused proof systems}
\label{ssec:focused}

The proof search approach to encoding Horn clause computation
results in the structuring of proofs with repeated
switching between a \emph{goal-reduction} phase and a
\emph{backchaining} phase~\cite{miller91apal}.
%
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
% \footnote{The terminology of negative and positive phases is a bit
%   unfortunate: historically, these terms do not refer to positive or negative
%   subformula occurrences but rather to certain semantic models used in
%   the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

@@

\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
\label{ssec:fpc}

\begin{footnotesize}
\begin{figure}
  \newcommand{\XXi}{}
  \newcommand{\bc}[2]{#1 >\!> #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi .}
\label{fig:augmented}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{A dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}
Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
% 
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

@@

>>>>

% {\color{red} The goal in this section is to say a bit about focusing
%   generally, and to give a bit of detail about FPC for Horn clauses.
%   Initial draft of text taken from \cite{blanco19ppdp}.
% }

\begin{metanote}
  (AM): Start with the logical reconstruction of \texttt{check}, to and then reverse engineer what we want/need to say about focusing and FPC. We \textbf{do} not talk about PBT yet, just logic programming (with proof terms). In sec 4 we package \texttt{check} in a PBT query and explain how \texttt{check} works for term generation.

  One way to proceed:

  \begin{itemize}
  \item One figure fig \ref{fig:interp} for Horn logic over Coq terms (no certificates) and its encoding with quotations 
  \item Then the logic behind the dependent kernel with certificates and proof terms
  \end{itemize}
\end{metanote}

\subsection{Horn logic in coq-elpi}

The proof search approach to encoding Horn clause computation
results in the structuring of proofs with repeated
switching between a \emph{goal-reduction} phase and a
\emph{backchaining} phase~\cite{miller91apal}.

In Fig.~\ref{fig:interp} we give a proof system for Horn logic and its encoding in coq-elpi
\todo{explain the interface, quotations etc}

\begin{footnotesize}
  \begin{figure}
  \newcommand{\XXi}{}
  \newcommand{\bc}[2]{#1 >\!> #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{A proof system for Horn logic and its encoding in coq-elpi .}
\label{fig:interp}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
%\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
%\caption{A dependent kernel for Coq terms.}
%\label{fig:kernel}
\end{figure}
\end{footnotesize}

\subsection{Focused proof systems with certificates}
\label{ssec:focused}

%
The notion of \emph{focused proof systems}~\cite{andreoli92jlc,liang09tcs}
generalizes this view of proof construction by identifying the
following two phases. % of proof construction.
\begin{enumerate}
\item The \emph{negative}%
% \footnote{The terminology of negative and positive phases is a bit
%   unfortunate: historically, these terms do not refer to positive or negative
%   subformula occurrences but rather to certain semantic models used in
%   the study of linear logic~\cite{girard87tcs}.}
%
  phase corresponds to  goal-reduction: in this phase,
  inference rules that involve \emph{don't-care-non\-de\-ter\-min\-ism} are
  applied.
%
  As a result, there is no need to consider backtracking over choices
  made in building this phase.

@@

\todo{Talk about the LJT thing: grammar for formulae and proof terms\cite{Herbelin94,LengrandDM06}?}

%\subsection{Certificate checking with expert predicates}
%\subsection{Adding certificates and experts to a proof system}
\label{ssec:fpc}

\begin{footnotesize}
\begin{figure}
    \newcommand{\XXi}{{\color{blue}{\Xi}}}

  \newcommand{\bc}[2]{#1 >\!> #2}
\[
\infer{\XXi\vdash t = t}
{%\eqExpert{\XXi}
}
\qquad
\infer{\XXi\vdash \true}
{%\trueExpert{\XXi}
}
\qquad
\infer{\XXi\vdash G_1\vee G_2}
{\XXi\vdash G_i%\qquad \orExpert{\XXi}{\XXi'}{i}
}
\qquad
\infer{\XXi\vdash \exists x. G}
{\XXi\vdash G[t/x]%\qquad \someExpert{\XXi}{\XXi'}{t}
}
\qquad 
\infer{\XXi\vdash G_1\wedge G_2}
{\XXi\vdash G_1\quad \XXi\vdash G_2%\quad \andExpert{\XXi}{\XXi_1}{\XXi_2}
}
\qquad
\infer{\XXi\vdash A}
      {\bc D A \quad D \in \Pscr}
\]
\vskip -18pt
\[
  \infer{\bc A A}{}{}\qquad
  \infer{\bc{\forall x.\, D} A}{\bc{D[t/x]} A}{}\qquad
  \infer{\bc{G\supset D} A}{\bc D A\quad \XXi\vdash G}{}
\]
\lstinputlisting[linerange={interp-end}]{code/dep-kernel.mod}
\caption{dep kernel.}
\label{fig:augmented}
%\lstinputlisting[linerange={sigs-end}]{code/dep-kernel.sig}
\lstinputlisting[linerange={check-end}]{code/dep-kernel.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}
\end{footnotesize}

Figure~\ref{fig:augmented} contains a simple proof system for Horn
clause provability in which each inference rule is augmented with an
additional premise involving an \emph{expert predicate}, a
certificate $\Xi$, and possibly continuations of certificates ($\Xi'$,
$\Xi_1$, $\Xi_2$) with extracted information from certificates (in the
case of $\vee$ and $\exists$).
%
%
% 
Although this proof system is a focused proof system, the
richness of focusing is not apparent in this simplified 
setting: thus, we drop the adjective ``focused'' from this point forward.

@@

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
